Algebraic Tori in Unitary Groups

### 2013-06-03 Paper.tex
Add in some more details of some arguments

\begin{proposition}
The discriminant of $H_{E,\sigma,\lambda}$ satisfies:
\[  (d,\delta_{A/K})_K =  \Cor_{E^\sigma/K}((\lambda (-1)^n f_z'(z), z)_{E^\sigma}) \cdot (\delta_{A/K},-1)_K^{n(n-1)/2}. \]
Moreover, the map:
\[ (E^\sigma)^\times/N_{E/E^\sigma}(E^\times) \rightarrow K^\times/N_{A/K}(A^\times) \]
given by:
\TODO{notation for hermitian discriminant}
\[ \lambda \mapsto (-1)^{n(n-1)/2} D\left( \tfrac{1}{2}\Tr_{E/A}\left( \frac{(-1)^n\lambda}{f'_z(z)} x\sigma(y) \right)\right) \]
is a homomorphism.
\end{proposition}
\begin{proof}
\TODO - complete sketch

@@

Provided $E/E^\sigma$ (equiv $A/K$) not split the formula should tell us non-trivial $d$ come from non-trivial $\lambda$.

Note that  $(E^\sigma)^\times/N_{E/E^\sigma}(E^\times)$ is $( K^\times/N_{A/K}(A^\times))^\ell$ where $\ell$ is the number of field factors of $E$.
\end{proof}

\begin{proposition}
The signature of $H_{E,\sigma,\lambda}$ at a real place of $K$ ramified in $A$ is:
\[ (\#\lambda^+,\#\lambda^-) \]
\end{proposition}

@@

\begin{theorem}
Given $H$ a Hermitian form and $(E,\sigma)$ an \'etale algebra with involution (of the second kind) over $A$.Then for all $p$ primes of $K$ there exists $\lambda_p \in E^\sigma$ with:
\[ H_p \simeq H_{E_p,\sigma,\lambda_p} \]
\TODO - I am pretty sure there are no conditions, I haven't fully checked.
\end{theorem}
\begin{proof}
\TODO - complete sketch

non-split Infinite places (use signature proposition)

non-split finite places (use discriminant map proposition)

Split places (trivial)
\end{proof}

\begin{theorem}
\TODO - phrasing

There exists $\lambda_p$ locally everywhere if and only if there exists $\lambda_p$ globally.
\end{theorem}
\begin{proof}
\TODO - complete sketch (also check that this is true).

Argument 1:

The local global obstruction allowing $\lambda$ such that $B_{E,\sigma,\lambda} \simeq B$ does not exist because for any place $p$ where $A/K$ is innert, then for every field factor of $E$ we have $E/E^\sigma$ is innert, thus there is no local global obstruction.

The equivalence of $B$'s and $H$'s should conclude the result.

\bigskip
Argument 2:

Optimistically global class field theory has a name for what we are doing.
We need a $\lambda$ with controlled number of positive/negative embeddings, asside from this we are trying to say that the map:
\[ (E^\sigma)^\times/N_{E/E^\sigma}(E^\times) \rightarrow K^\times/N_{A/K}(A^\times) \]
is surjective.
So in total what we want is surjective when we appropriately restrict to `totally postive' elements. This sounds like a very standard CFT type problem, and since by Argument 1 this result should hold... this result should have been proven by someone else before now.
\end{proof}

>>>>

\begin{proposition}\label{prop:discrimmap}
The discriminant of $H_{E,\sigma,\lambda}$ satisfies:
\[  (d,\delta_{A/K})_K =  \Cor_{E^\sigma/K}((\lambda (-1)^n f_z'(z), z)_{E^\sigma}) \cdot (\delta_{A/K},-1)_K^{n(n-1)/2}. \]
Moreover, the map:
\[ (E^\sigma)^\times/N_{E/E^\sigma}(E^\times) \rightarrow K^\times/N_{A/K}(A^\times) \]
given by:
\TODO{notation for hermitian discriminant}
\[ \lambda \mapsto (-1)^{n(n-1)/2} D\left( \tfrac{1}{2}\Tr_{E/A}\left( \frac{(-1)^n\lambda}{f'_z(z)} x\sigma(y) \right)\right) \]
is a homomorphism.
\end{proposition}
\begin{proof}
\TODO - complete sketch

@@

Provided $E/E^\sigma$ not split the formula should tell us non-trivial $d$ come from non-trivial $\lambda$.

Note that for a local field $(E^\sigma)^\times/N_{E/E^\sigma}(E^\times)$ is $( K^\times/N_{A/K}(A^\times))^\ell$ where $\ell$ is the number of field factors of $E$ which are $\sigma$-stable.
\end{proof}

\begin{proposition}\label{prop:discrimmapsurjective}
Let $K$ be $p$-adic fidle, $\RR$ or $\CC$ and fix $A$ a quadratic extension of $K$ (potentially split).
Fix $(E,\sigma)$ an \'etale algebra with involution (of the second kind) over $A$.
If $E$ is a field then the homomorphism of Proposition \ref{prop:discrimmap} is surjective.
\end{proposition}
\begin{proof}
The corestriction map is injective for local fields, and we are in the setting where both groups are isomorphic to $\{\pm1\}$, it follows immediately that the map is also surjective.
\end{proof}

\begin{proposition}\label{prop:signature}
Let $K$ be $\RR$ and $A$ be $\CC$.
If $E \simeq \CC$ then
The signature of $H_{E,\sigma,\lambda}$ is:
\[ (\#\lambda^+,\#\lambda^-) \]
If $E\simeq \CC\times\CC$ with $\sigma$ interchanging factors then the signature of $H_{E,\sigma,\lambda}$ is:
\[ (1,1). \]
\end{proposition}

@@

\begin{theorem}
Let $K$ be a $p$-adic field and fix $A$ a quadratic extension of $K$ (potentially split).
Given $H$ a Hermitian form and $(E,\sigma)$ an \'etale algebra with involution (of the second kind) over $A$. Then there exists $\lambda \in E^\sigma$ with:
\[ H \simeq H_{E,\sigma,\lambda} \]
if and only if there exists a factor $E_i^\sigma$ of $E^\sigma$ over which the corresponding factor of $E_i$ is not split.
\end{theorem}
\begin{proof}
If $A$ is split over $K$ then there is a unique Hermitian space and hence nothing to show.

In this setting the cohomology groups and the corestriction map of Propostion \ref{prop:discrimmap} are products of those appearing in Proposition \ref{prop:discrimmapsurjective}. The map of Propostion \ref{prop:discrimmap} is thus surjective if the cohomology groups are non-trivial. This is the case precisely if at least one factor of $E_i$ over $E_i^\sigma$ is not split.
\end{proof}

\begin{example}
Fix a prime numbers $p_1,p_2$ and $p_3$ such that $\left(\tfrac{p_1}{p_3}\right) = 1$ and $\left(\tfrac{p_2}{p_3}\right) = -1$.
Set $A=\QQ(\sqrt{p_2})$ with the unique non-trivial involution, and $E=\QQ(\sqrt{p_2},\sqrt{p_1p_2})$ with the unique involution extending $\sigma$ acting trivially on $\sqrt{p_1p_2}$.
Then $A_{p_3} = \QQ_{p_3}(\sqrt{p_2})$ and $E_{p_3} \simeq A_{p_3} \times A_{p_3}$ with $\sigma$ interchanging factors.
It follows that the torus $T_{E,\sigma}$ does not embed into any unitary group which is not quasi-split at $p_3$.
\end{example}

\begin{theorem}
Let $K$ be $\RR$ or $\CC$ and fix $A$ a quadratic extension of $K$ (potentially split).
Given $H$ a Hermitian form and $(E,\sigma)$ an \'etale algebra with involution (of the second kind) over $A$. Then there exists $\lambda \in E^\sigma$ with:
\[ H \simeq H_{E,\sigma,\lambda} \]
if and only if 
\TODO - phrase signature conditions
\end{theorem}
\begin{proof}
If $A$ is split over $K$ then there is a unique Hermitian space and hence nothing to show.
This is the case unless $A=\CC$ and $K=\RR$. In this case $E$ is $\CC^n$, however some of the factors might be interchanged by $\sigma$.
By Proposition \ref{prop:signature} the factors interchanged by $\sigma$ must contribute $(1,1)$ to the signature.
Wheras those not interchanged contribute $(1,0)$ or $(0,1)$ depending on the choice of $\lambda$.
From this we conclude the result.
\end{proof}

\begin{example}
\TODO - A CM-field field $A$ and a complex field $E$ such that $E\otimes_\QQ \RR$ has factors interchanged by an involution.
\end{example}

\begin{lemma}\label{Lem:liftsigma}
Let Let $K$ be a number field and fix $A$ a quadratic extension of $K$ (potentially split).
Fix $(E,\sigma)$ an \'etale algebra with involution (of the second kind) over $A$.
Write $E = E_s \times E_{ns}$ where $E_s \simeq E_s^\sigma \times E_s^\sigma$ and $\sigma$ interchanges factors and $E_{ns}$ is a product of $\sigma$-stable fields.
There exists $\tilde{\sigma} \in \Gal(\overline{K}/K)$ inducing $\sigma$ on $E_{ns}$ (that is a unique $\tilde{\sigma}$ induces the involution on each factor.
\end{lemma}
\begin{proof}
Write:
\[ E_{ns} = \prod E_i \]
and set $\tilde{E} = E_1^\sigma E_2^\sigma\ldots E_{n}^\sigma$ to be the composite field of all the fixed fields of the $E_i$.
Notice that $E_i \simeq E_i^\sigma\otimes_K A \simeq E_i^\sigma A$.
Now, $A$ and $\tilde{E}$ are algebraicly disjoint fields and for all $i$ there exists canonical embeddings of fields:
\[ \xymatrix{ E_i \simeq E_i^\sigma\otimes_K A \ar@{^(->}[r]\ar@{-}[d] &  \tilde{E} \otimes_K A \simeq \tilde{E}A \ar@{-}[d] \\ 
                      E_i^\sigma \ar@{^(->}[r]\ar@{-} & \tilde{E}.}
 \]
The field $\tilde{E}A$ has a unique involution $\tilde{\sigma}$ such that $(\tilde{E}A)^{\tilde{\sigma}} = \tilde{E}$. Since $E_i^{\tilde{\sigma}} = E_i^\sigma$ this involution restricts to $\sigma$ on the image of $E_i$.
Finally, $\tilde{\sigma}$ lifts to an element of $\Gal(\overline{K}/K)$, giving us the claim.
\end{proof}

\begin{theorem}
Let $K$ be a number field and fix $A$ a quadratic extension of $K$ (potentially split).
Given $H$ a Hermitian form and $(E,\sigma)$ an \'etale algebra with involution (of the second kind) over $A$. Then there exists $\lambda \in E^\sigma$ with:
\[ H \simeq H_{E,\sigma,\lambda} \]
if and only if there exists $\lambda_\fp$ for all localizations $\fp$ of $K$ (both archimedian and non-archimedian).
\end{theorem}
\begin{proof}
\TODO - complete sketch (also check that this is true).
\TODO - need to distinguish primes of $K$, $A$ and $E$.

Write $E^\sigma = \prod_{i=1}^n E^\sigma_i$.
We have the following diagram:
\[
\xymatrix{
(E^\sigma)^\times/N_{E/E^\sigma}(E^\times) \ar[r]^{(\cdot,z)} \ar[d]& H^2(E^\sigma, \pm 1) \ar[r] \ar[d]& \underset{E_{i,\fp} \text{ not split}}\oplus H^2(E^\sigma_{i,\fp},\pm 1) \ar[d]\\
K^\times/N_{A/K}(A^\times) \ar[r]^{(\cdot,\delta_{A/K})} & H^2(K, \pm 1) \ar[r] & \underset{A_\fp \text{ not split}}\oplus H^2(K_\fp,\pm 1)
}
\]
where the compositions of the horizontal maps are injective.
We must show that an element in $K^\times/N_{A/K}(A^\times) $ is in the image of the left verticle map, provided its image is in the image of the right hand verticle map.
We must do this in spite of the fact that the horizontal maps are not surjective.

The failure of surjectivity is captured by the following, the bottom map surjects onto the kernel of the map:
\[ \underset{A_\fp \text{ not split}}\oplus H^2(K_\fp,\pm 1)  \rightarrow \{ \pm1\} \]
given by $(x_\fp) \mapsto \prod x_\fp$. The top map surjects onto the kernel of the map:
\[ \underset{E_{i,\fp} \text{ not split}}\oplus H^2(E_{i,\fp}^\sigma,\pm 1)  \rightarrow \{ \pm1\}^n \]
given by $(x_{i,\fp}) \mapsto (\prod x_{1,\fp}, \ldots, \prod x_{n,\fp}) $.

To complete the argument, it suffices to show that the kernel of the map:
\[\underset{E_{i,\fp} \text{ not split}}\oplus H^2(E_{i,\fp}^\sigma,\pm 1)  \rightarrow \underset{A_\fp \text{ not split}}\oplus H^2(K_\fp,\pm 1) \]
surjects onto the kernel of the map:
\[ \{ \pm1\}^n \rightarrow \{ \pm1\}. \]
To show this, it suffices to do this for a single fixed $\fp$ a prime of $K$.
Fix $\fp$ any prime of $K$ where $\tilde{\sigma}$ of Lemma \ref{Lem:liftsigma} acts as Frobenius. Then for all $i$ there exists $\fp_i | \fp$ prime of $E_i^\sigma$ for which $E_{\fp_i,i}$ is non-split. It follows that $\oplus_i H^2(E_{\fp_i,i}^\sigma,\pm1)$ surjects onto $\{\pm 1\}^n$, thus this also holds for the larger group, and hence also for the kernels.
\end{proof}

\begin{theorem}
\TODO - summarize a pure global existance statement.
\end{theorem}

\begin{remark}
\TODO - Can phrase this as: The reflex algebra splits the even clifford algebra + signature conditions.
Is there a meaningfull interpretation of the clifford algebra in this context?
Can we interpret it as telling us about a quaternioninc symplectic group embedding of the unitary group?
\end{remark}
